Estimating the number of clusters using Cross Validation

### 2015-08-20 cvclust.tex
move counterexample to main text

@@

\section{Asymptotic Analysis of Gabriel Cross-Validation}

@@

Here, we show that Gabriel Cross-Validation is consistent when the data come
from a single cluster; the algorithm picks $k = 1$ with probability tending to
$1$ as the number of rows in the data matrix $\dataX$ increase.

To prove consistency of Gabriel cross-validation, it suffices to analyze a
single fold of the cross-validation procedure.  We will assume the following
setup:

\begin{theorem}
For each $n$, let $m(n)$ be a nonnegative integer with $m(n) \to \infty$ as $n
\to \infty$, and let $\{ (X_{n,i}, Y_{n,i}) \}_{i=1}^{n+m(n)}$ be independent
identically distributed random variables.  For any fixed $k$, let $\CV(k)$
be the cross-validation error from a single Gabriel CV fold, computed as
described in Sec.~\ref{sec:gabriel-cv-algorithm}.  Suppose that
\begin{enumerate}[label=(\roman*)]
\item the distribution for each $(X, Y)$ pair is such that $X$ and $Y$ have
  marginal distributions $P^X$ and $P^Y$, and the vectors $X$ and $Y$ are
  independent of each other;

\item the distribution $P^Y$ satisfies $\int \|y\|^2 P^Y(dy) < \infty$;

\item the support of $P^Y$ contains at least two points.
\end{enumerate}
Then for any fixed $k > 1$, almost surely eventually  $\CV(k) > \CV(1)$.
\end{theorem}
\begin{proof}
Set $\muY = \E(Y)$ and $\SigmaY = \cov(Y)$.  The strong law of large
numbers implies that
\(
  \CV(1) \toas \int \|y - \mu^Y\|^2 P(dy) = \tr(\SigmaY).
\)

Now suppose that $k > 1$.  Let $\bmuY_1, \dotsc, \bmuY_k$ be the cluster
centers gotten from applying $k$-means to $Y_{n,1}, \dotsc, Y_{n,n}$.  Since
the support of $P^Y$ has at least two points,
Lemma~\ref{lem:kmeans-existence}, implies that almost surely eventually at
least two of the clusters will be nonempty and have distinct centers.  Without
loss of generality, assume that $k$ is less than or equal to the cardinality
of the support of $P^Y$.  Theorem~\ref{thm:kmeans-consistency} and
Lemma~\ref{lem:kmeans-separation} imply that there exists an $r > 0$ such that
almost surely eventually at most one cluster center is within distance $r$ of
$\muY$.  Without loss of generality suppose that $\bmuY_1$ is the closest
cluster center to $\muY$.

Let $\bmuX_1, \dotsc, \bmuX_k$ be the corresponding sample means computed from
$X_{n,1}, \dotsc, X_{n,n}$, with the $i$th observation assigned to cluster
$\hGY_{n,i} = \argmin_g \| Y_{n,i} - \bmuY_{n,i} \|^2.$  Consider the
classification rule $\hGY(X)$, which assigns $X$ to the class $g$ minimizing
$\|X - \bmuX_g\|^2$, breaking ties uniformly at random between classes of
equal distance.

[TODO: finish]

\end{proof}

\subsection{Well-Separated Clusters with Noise}

@@

\section{Technical Lemmas}

@@

\section{Counterexample with correlated noise}

Suppose that $(X, Y)$ are jointly mean-zero multivariate normal with unit
marginal variances and correlation $\rho$.  In this case, the data are drawn
from a single cluster; the true number of clusters is~$1$.  We will show that
Gabriel cross-validation is inconsistent if $\rho > 0.5$, in the sense that $k
= 2$ will be preferred to $k = 1$ almost surely for large enough $n$.

Given a sample $(X_1, Y_1), \dotsc, (X_n, Y_n)$, we first apply $k$-means to
$\{ Y_i \}_{i=1}^{n}$.  If $k = 1$ and $n$ is large enough, then the centroid
will be close to $\E(Y) = 0$, in the sense that the centroid converges almost
surely to this value as $n$ tends to infinity \citep{pollard1981strong}.
Similarly, if $k = 2$ and $n$ is large enough, then the centroids will tend to
$\E(Y \mid Y > 0) = \sqrt{2/\pi}$ and $\E(Y \mid Y < 0) = -\sqrt{2/\pi}$.  We
have used Lemma~\ref{lem:truncated-normal-moments} to compute the
expectations.  In $k = 2$ case, then if $n$ is large enough, the cluster
assignment will be determined according to whether $Y > 0$.

In the $k = 1$ case, if $n$ is large enough, the average Gabriel
cross-validation error converges to $\E(Y^2) = 1$. In the $k = 2$ case, if $n$
is large enough and $\rho > 0$, then the linear decision rule for the $X$
variables will be determined according to whether $X > 0$; if $\rho < 0$ then
the decision is according to whether $X < 0$.  In the $\rho > 0$ case, the
error converges to $\E[(Y - a)^2 1\{ X > 0\}] + \E[(Y + a)^2 1\{X < 0\}]$,
where $a = \sqrt{2/\pi}$.  From the joint normality of $X$ and $Y$, it follows
that $Y \mid X$ is normal with mean $\rho X$ and variance $(1 - \rho^2)$, so
that $\E[(Y - a)^2 \mid X] = (\rho X - a)^2 + (1 - \rho^2)$.  Applying
Lemma~\ref{lem:truncated-normal-moments}, in the $k = 2$ case we get that the
Gabriel cross-validation error converges to $1 + a^2 (1 - 2 \rho)$.  In the
$\rho < 0$ case, it converges to $1 + a^2 (1 + 2 \rho)$.  In particular, if
$|\rho| > 0.5$, then the asymptotic cross-validation error for $k = 2$ will be
smaller than for $k = 1$.

We confirm this with a simulation.  We perform $10$ replicates.  In each
replicate, we generate $20000$ observations from a mean-zero bivariate normal
distribution with unit marginal variances and correlation $\rho$.  We perform
a single $2 \times 2$ fold of Gabriel cross-validation and report the
cross-validation mean squared error for the number of clusters $k$ ranging
from $1$ to $5$.  Figure~\ref{fig:nullcorr-equal} shows the cross-validation
errors for all $10$ replicates.  The simulation demonstrates that in the
Gabriel cross-validation criterion chooses the correct answer $k = 1$ whenever
$\rho < 0.5$; the criterion chooses $k = 2$ clusters whenever $|\rho| > 0.5$.

>>>>

\newcommand{\OhP}{O_p}

@@

\section{Analysis of Gabriel Cross-Validation with Gaussian Noise}

@@

Now we we analyze the asymptotic performance of Gabriel Cross-Validation, in
the case of Gaussian noise. Our main result is that with single-cluster
Gaussian data, if the predictor and
response columns of $\dataX$ are weakly correlated or independent, then the
method will correctly choose to $k = 1$ cluster. We state the result in the
case where $\dataX$ has 2 columns, but it is straightforward to generalize
this result to higher dimensions.

\begin{proposition}
Suppose that $\{ (X_i, Y_i) \}_{i=1}^{n + m}$ is that data from a single fold
of Gabriel cross-validation, where each $(X,Y)$ pair in $\R^2$ is an
independent draw from a mean-zero multivariate normal distribution with unit
marginal variances and correlation $\rho$.  In this case, the data are drawn
from a single cluster; the true number of clusters is~$1$.  If if $\rho <
0.5$, then $k = 1$ will be preferred to $k = 2$ for large enough $n$ and $m$.
\end{proposition}

\begin{proof}(Outline.)

Given $(X_1, Y_1), \dotsc, (X_n, Y_n)$, we first apply $k$-means to $\{ Y_i
\}_{i=1}^{n}$. With $k = 1$, the single-cluster centroid will
be equal to $\bar Y_n$, the sample mean of the $Y_1, \dotsc, Y_n$, approximately equal to
$\E(Y) = 0$, with error of size $\OhP(n^{-1/2})$. The cross-validation error
will be
\[
  \CV(1) = \frac{1}{m} \sum_{i=n+1}^{n+m} \| Y_i - \bar Y_n \|^2
         = 1 + \OhP(m^{-1/2}) + \OhP(n^{-1/2}).
\]

Now we will consider the $k = 2$ case.  If $n$ is large enough, then
\citet{pollard1981strong} showed that the centroids $\bmuY_1$ and $\bmuY_2$
will be close to $\E(Y \mid Y > 0) = \sqrt{2/\pi}$ and $\E(Y \mid Y < 0) =
-\sqrt{2/\pi}$.  We have used Lemma~\ref{lem:truncated-normal-moments}
(Appendix~\ref{app:technical-lemmas}) to compute the expectations.  Further,
\citet{pollardclt} showed that the errors will be of size $\OhP(n^{-1/2})$.

If $\rho > 0$ and $n$ is large enough, then classification rule learned from
$\{(X_i, \hGY_i)\}_{i=1}^{n}$ variables will be determined according to
whether $X > 0$; if $\rho < 0$ then the decision is according to whether $X <
0$.  More specifically, the decision boundary will be at $0 + \OhP(n^{-1/2})$.

In the $\rho > 0$ case, the cross-validation error will be
\begin{align*}
  \CV(2)
  &=
    \frac{1}{m}
    \sum_{i=n+1}^{n+m}
      \| (Y_i - \bmuY_1) 1\{\hGX_i = 1\} \|^2
      +
      \| (Y_i - \bmuY_2) 1\{\hGX_i = 2\} \|^2
\\
  &=
    \E[(Y - a)^2 1\{ X > 0\}] + \E[(Y + a)^2 1\{X < 0\}]
      + \OhP(m^{-1/2}) + \OhP(n^{-1/2}),
\end{align*}
where $a = \sqrt{2/\pi}$.  From the joint normality of $X$ and $Y$, it follows
that $Y \mid X$ is normal with mean $\rho X$ and variance $(1 - \rho^2)$, so
that $\E[(Y - a)^2 \mid X] = (\rho X - a)^2 + (1 - \rho^2)$.  Applying
Lemma~\ref{lem:truncated-normal-moments}, we get that for large $m$ and $n$,
the Gabriel cross-validation error is close to
$1 + a^2 (1 - 2 \rho)$.

In the $\rho < 0$ case, a similar calculation shows that $\CV(2)$ is close to
$1 + a^2 (1 + 2 \rho)$.  In particular, if $|\rho| < 0.5$, then with
probability tending to $1$ and $m$ and $n$ increase, the asymptotic
cross-validation error for $k = 1$ will be smaller than for $k = 2$.
\end{proof}

We confirm this result with a simulation.  We perform $10$ replicates.  In each
replicate, we generate $20000$ observations from a mean-zero bivariate normal
distribution with unit marginal variances and correlation $\rho$.  We perform
a single $2 \times 2$ fold of Gabriel cross-validation and report the
cross-validation mean squared error for the number of clusters $k$ ranging
from $1$ to $5$.  Figure~\ref{fig:nullcorr-equal} shows the cross-validation
errors for all $10$ replicates.  The simulation demonstrates that in the
Gabriel cross-validation criterion chooses the correct answer $k = 1$ whenever
$\rho < 0.5$; the criterion chooses $k = 2$ clusters whenever $|\rho| > 0.5$.

\subsection{Two Clusters}

@@

\section{Technical Lemmas}
\label{app:technical-lemmas}

@@
